---
title: "nmf attempt 2"
output: html_document
date: "2025-02-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(dplyr)
library(NMF)
library(tidyverse)
setwd("//wsl.localhost/Ubuntu/home/ideluis42/nmf-air-quality")
library(ggplot2)
library(tidyr)
library(dplyr)
library(scales)  # For percentage formatting
library(openair)
```

```{r load data}
load("cleaned_modpm.RData")
```

```{r get random seed of data}
set.seed(123)
X_sample<-aq_df_filtered[sample(nrow(aq_df_filtered), 5000), ]
X_sample[X_sample < 0] <- 0  # Replace negative values with 0
X_sample <- na.omit(X_sample)  # Remove rows with NA values

```


```{r}
# remove non numeric data -- can't put non numeric data through nmf
numeric_cols <- c("bin0", "bin1", "bin2", "bin3", "bin4", "bin5", "bin6", "bin7",
                  "bin8", "bin9", "bin10", "bin11", "bin12", "bin13", "bin14", "bin15",
                  "bin16", "bin17", "bin18", "bin19", "bin20", "bin21", "bin22", "bin23",
                  "pm1", "pm25", "pm10")

X <- X_sample[, numeric_cols]
X <- as.matrix(X) # transform into matrix

```

```{r}
summary(X)
```

```{r}
# .options = t stores the residuals track
res <- nmf(X, 4, method="brunet", seed=123, .options = "t")

```

```{r}
summary(res)
```

```{r}
plot(res)
# Check how well the data is approximated
heatmap(basis(res) %*% coef(res))
```

# Heat Map Notes:

-   The labels along the X-axis are the features used in the NMF analysis

-   The Y-axis represents sample

-   The clustering dendrogram (tree-like structure) suggests that similar samples are grouped together

-   Dark red regions indicate higher values.

-   Light yellow regions indicate lower values (almost zero)

-   This suggests that some features (e.g., pm10, bin3) have higher values in certain groups of samples

-   The features are not evenly distributed; most values are close to zero except for a few high-intensity ones

-   Only certain variables (e.g., pm10, bin3, bin19) seem to contribute significantly to the structure.

    -   The clusters may indicate groups of similar environmental conditions (e.g., high vs. low pollution levels).


```{r}
## first need to access factor data
basis_df <-
  data.frame(basis(res))

coef_df <-
  data.frame(coef(res))


```


```{r}

## want to create a bar plot, where each bin is a bar, separated by the 4 factors
# first need to create a factor column

coef_df$Factor <- as.factor(1:nrow(coef_df))

coef_df <- coef_df |>
  select(Factor, everything())

# Reshape Data to Long Format
coef_df_long <- coef_df |>
  pivot_longer(cols = -Factor, names_to = "Bin", values_to = "Value")

glimpse(coef_df_long)
# Set threshold to filter out small values
threshold <- 1e-5  # Arbitrary

coef_df_long <- coef_df_long |>
  filter(Value > threshold)  # Keep only bins above threshold

# Plot the Data
ggplot(coef_df_long, aes(x = Bin, y = Value, fill = Bin)) +
  geom_bar(stat = "identity", position = "dodge") +  # Grouped bars
  facet_wrap(~Factor, scales = "free") +  # Separate by factor
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_d()  # Improve color scheme
```
```{r}
# attempt at time series plot
basis_df <-
  basis_df |>
    mutate(timestamp = X_sample$timestamp,
           date = X_sample$date)

basis_df <-
  basis_df |>
    select(timestamp, everything())

timePlot(basis_df, pollutant = c("X1", "X2", "X3", "X4"), group = TRUE) # time plot ugly af


my_plot <- timeVariation(basis_df, pollutant = c("X1", "X2", "X3", "X4"), ci = FALSE) 
plot(my_plot, subset = "hour")


```



