---
title: "NMF Documentation"
author: "Isabel de Luis"
date: "2025-04-09"
output: html_document
---

# Non-Negative Matrix Factorization Pipeline for Air Quality Analysis

By Isa de Luis and Gia-Uyen Tran

## Introduction

{Isa doesn't want to write this rn}

## Setup

The first you need to do in order to run this pipeline is setup your R so that it can access the data files. Set the `root_dir` variable below to the folder that the pipeline is in. Within that folder, you should create a `data` folder that you can put your `Rdata` files of air quality data.

You can download Air Partners air quality data here: please note that if you change any file names, you will need to change the code appropriately so that it works.

### Isa Side Bar: Might be able to create pipeline such that all they need to do is download a google drive folder and BAM everything works

```{r setup, include=FALSE}

root_dir <-  NA # Change this to be the path to the folder this file is in
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "////wsl.localhost/Ubuntu/home/ideluis42/nmf-air-quality/")
```

Now, you need to load the following packages that this file uses to do the analysis.

```{r load packages}

# if any packages are not installed on machine, install them
if (!require(dplyr)) install.packages("dpylr")
if (!require(NMF)) install.packages("NMF")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(tidyr)) install.packages("tidyr")
if (!require(scales)) install.packages("scales")
if (!require(openair)) install.packages("openair")

library(dplyr)
library(NMF)
library(tidyverse)
library(ggplot2)
library(tidyr)
library(scales)  # For percentage formatting
library(openair)
```

## Loading Data
Now that your R is setup, you can start your analysis! The first thing you want to do is make sure that your data is in the `/data` folder that you created earlier. Set the `data_path` variable below to the path to the data. If you've followed the instructions above, it should be something like `data/filename.RData`.

```{r data load}
 #isa to do make it less specific to your machine
load(data_path)
```

If you're using Air Partners data, you should now have a data frame in your environment called something like `aq_df_filtered`. Double click on it to open it open and inspect the variables. Since we want to analyze air quality data, we only want to run NMF on our air pollutant columns. Each data set has a slightly different set of pollutants, so you need to look at your dataset and determine what variables you want to run the NMF analysis on. You can indicate what columns you want the analysis run on in the `cols_analysed` variable below. For our example dataset from South Bronx, we have the following air quality columns:
  - PM1
  - PM2.5
  - PM10
  - Bins 0 - 17
  - CO
  - NO
  - NO2
  - O3

```{r}
cols_analysed <- c("pm1", 
                   "pm25", 
                   "pm10", 
                   "co", 
                   "no", 
                   "no2", 
                   "o3",
                   "bin0",
                   "bin1",
                   "bin2",
                   "bin3",
                   "bin4",
                   "bin5",
                   "bin6",
                   "bin7",
                   "bin8",
                   "bin9",
                   "bin10",
                   "bin11",
                   "bin12",
                   "bin13",
                   "bin14",
                   "bin15",
                   "bin16",
                   "bin17") # columns to be analysed, change with what's in each unique dataset
```

## Helper Functions
To help facilate our analysis, we've written a series of helper functions! All you have to do is run them on the dataset.

The first helper function is `find_rank`. This function helps you find the right number of factors for your analysis by computing the coephentic correlation for NMF analysis with different numbers of factors. The higher the copehentic correlation, the better your results are (? hate how this is written also not sure if true). Find rank takes three parameters: {isa to do}

The next helper function is `run_nmf`, which does exactly what its name implies -- it runs the NMF analysis on a matrix. 

```{r helper functions}
find_rank <- function(X, start_k, end_k) {
  
  # factors
  k_values <- start_k:end_k
  
  coph_cor_values <- numeric(length(k_values))
  
  for (i in seq_along(k_values)) {
    k <- k_values[i]
    
    # Perform NMF with multiple runs (50 for stability)
    nmf_result <- nmf(X, rank = k, nrun = 50, method = "brunet")
    
    # Extract the consensus matrix
    consensus_mat <- consensus(nmf_result)
    
    # Compute cophenetic correlation
    coph_cor_values[i] <- cophcor(consensus_mat, linkage = "average")
    
  }
  
  # Display cophenetic correlation values for each k
  df_x <- data.frame(k = k_values, cophenetic_correlation = coph_cor_values)
  
  df_x |>
      ggplot(aes(x = k, y = cophenetic_correlation)) +
      geom_point()
  }


run_nmf <- function(df, cols, rank) {
  X_sample <- na.omit(df)  # Remove rows with NA values

  X_sample <- X_sample[sample(nrow(X_sample), 5000), ]
  X_sample[X_sample < 0] <- 0  # Replace negative values with 0
  
  X <- X_sample[, cols]
  X <- as.matrix(X) # transform into matrix
  
  res <- nmf(X, rank, method = "brunet", seed=123, .options = "t")
  
  return(res)
}

generate_bar_plot <- function(coef_df) {
  coef_df$Factor <- as.factor(1:nrow(coef_df))

  coef_df <- coef_df |>
    select(Factor, everything())
  
  # Reshape Data to Long Format
  coef_df_long <- coef_df |>
    pivot_longer(cols = -Factor, names_to = "Bin", values_to = "Value")
  
  glimpse(coef_df_long)
  # Set threshold to filter out small values
  threshold <- 1e-4  # Arbitrary
  
  coef_df_long <- coef_df_long |>
    filter(Value > threshold)  # Keep only bins above threshold
  
  # Plot the Data
  plot <- ggplot(coef_df_long, aes(x = Bin, y = Value, fill = Bin)) +
    geom_bar(stat = "identity", position = "dodge") +  # Grouped bars
    facet_wrap(~Factor, scales = "free") +  # Separate by factor
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_viridis_d()  # Improve color scheme
  
  return(plot)
  
}

generate_time_variation <- function(basis_df) {
  basis_df <-
    basis_df |>
      mutate(timestamp = X_sample$timestamp_local.x,
             date = X_sample$date)

  basis_df <-
    basis_df |>
      select(timestamp, everything())

  
  my_plot <- timeVariation(basis_df, pollutant = c("X1", "X2", "X3"), ci = FALSE, plot = FALSE) 
  
  return(my_plot)
}

generate_polar_plot <- function(basis_df, rank) {
  basis_df <-
  basis_df |>
    mutate(wd = X_sample$wd,
           ws = X_sample$ws,
           lat = X_sample$lat,
           lon = X_sample$lon)
  
  for (i in 1:rank) {
    pol = paste0("X", i)
    polarPlot(basis_df, pollutant = pol)
  }
}

```

## Network Analysis

### NMF Analysis

```{r nmf analysis}
X_sample <- na.omit(aq_df_filtered)  # Remove rows with NA values

X_sample <- X_sample[sample(nrow(X_sample), 100000), ]
X_sample[X_sample < 0] <- 0  # Replace negative values with 0

X <- X_sample[, cols_analysed]
X <- as.matrix(X) # transform into matrix
```

#### Rank Factorization

```{r}
find_rank(X, 2, 6)
```

```{r}
res <- nmf(X, rank = 4, method = "brunet", seed=123, .options = "t")

basis_df <-
  data.frame(basis(res))

coef_df <-
  data.frame(coef(res))
```

```{r plots}
generate_bar_plot(coef_df)
generate_time_variation(basis_df)
generate_polar_plot(basis_df, 5)
```

## By Sensor Analysis

### NMF Analysis

```{r get sensors}
sensors <- unique(aq_df_filtered$sn)

sensors
```

```{r by sensor nmf analysis}

for (sensor in sensors){
  df <- aq_df_filtered |>
    filter(sn == sensor)
  
  res <- run_nmf(df, cols_analysed, 3)
  
  basis_df_name <- paste0("basis_df_", sensor)
  coef_df_name <- paste0("coef_df_", sensor)
  
  basis_df <-
    data.frame(basis(res))

  coef_df <-
    data.frame(coef(res))
  
  assign(basis_df_name, basis_df)
  assign(coef_df_name, coef_df)
}

```

### Bar Plots

```{r sensor bar plots}

for (sensor in sensors) {
  coef_df_name <- paste0("coef_df_", sensor)
  coef_df <- get(coef_df_name)
  my_plot <- generate_bar_plot(coef_df)
  
  plot(my_plot) +
    ggtitle(paste0(sensor), " factors bar plot")
  
  plot_name <- paste0("bar_plot_", sensor)
  
  assign(plot_name, plot)
}

```

### Time Variation Plots

```{r}
# generate time variation plot

for (sensor in sensors) {
  basis_df_name <- paste0("basis_df_", sensor)
  basis_df <- get(basis_df_name)
  my_plot <- generate_time_variation(basis_df)
  
  plot(my_plot)
  
  plot_name <- paste0("time_variation_", sensor)
  
  assign(plot_name, plot)
}
```

### Polar Plots

```{r}
for (sensor in sensors) {
  basis_df_name <- paste0("basis_df_", sensor)
  basis_df <- get(basis_df_name)
  generate_polar_plot(basis_df, 3)
}
```
